---
title: "null_visualizations_MC3"
author: "m quigg"
date: "2026-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Viola CDmetaPOP Simulation Visualization
This is to visualize the 3 MC replicates of the null model.  

## Set up
### Working Driectory and Library
```{r, include=FALSE}
setwd("C:/Users/mquig/Desktop/DED/LandGen_viola_sims")

library(tidyverse)
library(tools)
library(utils)
```

### Import Files
```{r}
#### MC1 ####
data_null_pop_MC1 = read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc0species0/summary_popAllTime.csv", stringsAsFactors = FALSE, sep = ",")

data_null_class_MC1=read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc0species0/summary_classAllTime.csv", stringsAsFactors = FALSE, sep = ",")


#### MC2 ####
data_null_pop_MC2 = read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc1species0/summary_popAllTime.csv", stringsAsFactors = FALSE, sep = ",")

data_null_class_MC2=read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc1species0/summary_classAllTime.csv", stringsAsFactors = FALSE, sep = ",")


#### MC3 ####
data_null_pop_MC3 = read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc2species0/summary_popAllTime.csv", stringsAsFactors = FALSE, sep = ",")

data_null_class_MC3=read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc2species0/summary_classAllTime.csv", stringsAsFactors = FALSE, sep = ",")
```

## Population Over Time
The goal is to make a graph that depicts the total population across all patches over time, for each of the MCs. It is a line graph with 3 lines, one for each replicate.
```{r}
#### Extract the data ####

### MC1 ###
pop_size_MC1=data_null_pop_MC1 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("MC1", paste0("patch_", 1:9)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("patch_"), as.numeric)) %>% 
  select(Year, MC1)

### MC2 ###
pop_size_MC2=data_null_pop_MC2 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("MC2", paste0("patch_", 1:9)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("patch_"), as.numeric)) %>% 
  select(Year, MC2)

### MC3 ###
pop_size_MC3=data_null_pop_MC3 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("MC3", paste0("patch_", 1:9)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("patch_"), as.numeric)) %>% 
  select(Year, MC3)


#### Make new dataframe ####
###combine dataframes
pop_size_MC=merge(merge(pop_size_MC1, pop_size_MC2, by="Year"), pop_size_MC3, by="Year")

###Convert to long format for graphing
pop_size_long=tidyr::pivot_longer(pop_size_MC, cols = starts_with("MC"), names_to = "MC", values_to = "Population Size")

#### graph it! ####
ggplot(pop_size_long, aes(x = Year, y = as.numeric(`Population Size`), color = MC)) +
  geom_line(linewidth = 1, alpha = 0.6) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "MC") +
  ggtitle("Total Population")
```

## Population Size by Cohort
The goal here is to create 3 graphs (one per MC) that show the distribution of population size based on age classes.
```{r}
#### Extract the age data ####
### MC 1###
age_pop_MC1=data_null_class_MC1 %>% 
  select(Year, N_Initial_Class) %>% 
  separate(N_Initial_Class, into = c("MC1", paste0("age_", 0:24)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("age_"), as.numeric)) %>% 
  select(-MC1)

### MC 2###
age_pop_MC2=data_null_class_MC2 %>% 
  select(Year, N_Initial_Class) %>% 
  separate(N_Initial_Class, into = c("MC2", paste0("age_", 0:24)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("age_"), as.numeric)) %>% 
  select(-MC2)

### MC 3###
age_pop_MC3=data_null_class_MC3 %>% 
  select(Year, N_Initial_Class) %>% 
  separate(N_Initial_Class, into = c("MC3", paste0("age_", 0:24)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("age_"), as.numeric)) %>% 
  select(-MC3)

#### convert to long format ####
### MC1 ###
##pivot
long_age_MC1=tidyr::pivot_longer(age_pop_MC1, cols = starts_with("age_"), names_to = "Age", values_to = "Population")
##clean up columns
cohort_age_MC1=long_age_MC1 %>% 
  separate(Age, into = c("delete", "Age"), sep = "_") %>% 
  select(-delete)
cohort_age_MC1$Age=as.numeric(cohort_age_MC1$Age)


### MC2 ###
##pivot
long_age_MC2=tidyr::pivot_longer(age_pop_MC2, cols = starts_with("age_"), names_to = "Age", values_to = "Population")
##clean up columns
cohort_age_MC2=long_age_MC2 %>% 
  separate(Age, into = c("delete", "Age"), sep = "_") %>% 
  select(-delete) 
cohort_age_MC2$Age=as.numeric(cohort_age_MC2$Age)

### MC3 ###
##pivot
long_age_MC3=tidyr::pivot_longer(age_pop_MC3, cols = starts_with("age_"), names_to = "Age", values_to = "Population")
##clean up columns
cohort_age_MC3=long_age_MC3 %>% 
  separate(Age, into = c("delete", "Age"), sep = "_") %>% 
  select(-delete)
cohort_age_MC3$Age=as.numeric(cohort_age_MC3$Age)


#### group into cohorts ####
### MC1 ###
##organize into classes
cohort_MC1=cohort_age_MC1 %>% 
  mutate(
    cohort = case_when(
      Age <=5 ~ "Ages 0-5",
      Age >= 6 & Age <=10 ~ "Ages 6-10",
      Age >= 11 & Age <=15 ~ "Ages 11-15",
      Age >= 16 & Age <=20 ~ "Ages 16-20",
      Age >= 21 & Age <=25 ~ "Ages 21-25")) %>% 
  na.omit() %>% 
  select(-Age)
##summarize populations in classes
cohort_MC1_summary=cohort_MC1 %>% 
  group_by(Year, cohort) %>% 
  summarise(population=sum(Population))

### MC1 ###
##organize into classes
cohort_MC2=cohort_age_MC2 %>% 
  mutate(
    cohort = case_when(
      Age <=5 ~ "Ages 0-5",
      Age >= 6 & Age <=10 ~ "Ages 6-10",
      Age >= 11 & Age <=15 ~ "Ages 11-15",
      Age >= 16 & Age <=20 ~ "Ages 16-20",
      Age >= 21 & Age <=25 ~ "Ages 21-25")) %>% 
  na.omit() %>% 
  select(-Age)
##summarize populations in classes
cohort_MC2_summary=cohort_MC2 %>% 
  group_by(Year, cohort) %>% 
  summarise(population=sum(Population))

### MC1 ###
##organize into classes
cohort_MC3=cohort_age_MC3 %>% 
  mutate(
    cohort = case_when(
      Age <=5 ~ "Ages 0-5",
      Age >= 6 & Age <=10 ~ "Ages 6-10",
      Age >= 11 & Age <=15 ~ "Ages 11-15",
      Age >= 16 & Age <=20 ~ "Ages 16-20",
      Age >= 21 & Age <=25 ~ "Ages 21-25")) %>% 
  na.omit() %>% 
  select(-Age)
##summarize populations in classes
cohort_MC3_summary=cohort_MC3 %>% 
  group_by(Year, cohort) %>% 
  summarise(population=sum(Population))

#### graph it ####
### MC1 ###
ggplot(cohort_MC1_summary, aes(x = Year, y = population, color = cohort)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "Cohort") +
  ggtitle("MC1")

### MC2 ###
ggplot(cohort_MC2_summary, aes(x = Year, y = population, color = cohort)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "Cohort") +
  ggtitle("MC2")

### MC3 ###
ggplot(cohort_MC3_summary, aes(x = Year, y = population, color = cohort)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "Cohort") +
  ggtitle("MC3")
```
## Population by Patch
The goal here is to create 3 graphs (one for each MC) depicting the population by patch over time.
```{r}
#### MC1 ####
###extract pop data by patch
pop_patch_MC1=data_null_pop_MC1 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("patch_"), as.numeric))
 
###pivot
long_pop_patch_MC1=tidyr::pivot_longer(pop_patch_MC1, cols = starts_with("patch_"), names_to = "Patch", values_to = "Population")

###graph it
ggplot(long_pop_patch_MC1, aes(x = Year, y = Population, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population Size", color = "Patch") +
  ggtitle("MC1") 
  #+ geom_vline(xintercept = 25, color = "black", linetype = "dashed")
  
#### MC2 ####
###extract pop data by patch
pop_patch_MC2=data_null_pop_MC2 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("patch_"), as.numeric))
 
###pivot
long_pop_patch_MC2=tidyr::pivot_longer(pop_patch_MC2, cols = starts_with("patch_"), names_to = "Patch", values_to = "Population")

###graph it
ggplot(long_pop_patch_MC2, aes(x = Year, y = Population, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population Size", color = "Patch") +
  ggtitle("MC2")   
  #+ geom_vline(xintercept = 25, color = "black", linetype = "dashed")
  
#### MC3 ####
###extract pop data by patch
pop_patch_MC3=data_null_pop_MC3 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("patch_"), as.numeric))
 
###pivot
long_pop_patch_MC3=tidyr::pivot_longer(pop_patch_MC3, cols = starts_with("patch_"), names_to = "Patch", values_to = "Population")

###graph it
ggplot(long_pop_patch_MC3, aes(x = Year, y = Population, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population Size", color = "Patch") +
  ggtitle("MC3")   
  #+ geom_vline(xintercept = 25, color = "black", linetype = "dashed")
```
## Heterozygosity
I want one graph with observed and expected heterozygosity for each of the MC replicates.
```{r}
#### extract data ####
### MC1 ###
het_MC1=data_null_pop_MC1 %>% 
  select(Year, Ho, He) %>% 
  separate(Ho, into = c("Ho_MC1", paste0("Ho_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  separate(He, into = c("He_MC1", paste0("He_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("Ho"), as.numeric)) %>% 
  mutate(across(starts_with("He"), as.numeric)) %>% 
  select(Year, Ho_MC1, He_MC1)

### MC2 ###
het_MC2=data_null_pop_MC2 %>% 
  select(Year, Ho, He) %>% 
  separate(Ho, into = c("Ho_MC2", paste0("Ho_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  separate(He, into = c("He_MC2", paste0("He_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("Ho"), as.numeric)) %>% 
  mutate(across(starts_with("He"), as.numeric)) %>% 
  select(Year, Ho_MC2, He_MC2)

### MC3 ###
het_MC3=data_null_pop_MC3 %>% 
  select(Year, Ho, He) %>% 
  separate(Ho, into = c("Ho_MC3", paste0("Ho_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  separate(He, into = c("He_MC3", paste0("He_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("Ho"), as.numeric)) %>% 
  mutate(across(starts_with("He"), as.numeric)) %>% 
  select(Year, Ho_MC3, He_MC3)

#### Make new dataframe ####
###combine dataframes
het_MC=merge(merge(het_MC1, het_MC2, by="Year"), het_MC3, by="Year")

###Convert to long format for graphing
het_long=tidyr::pivot_longer(het_MC, cols = starts_with("H"), names_to = "Replicate", values_to = "Heterozygosity")

#### graph it! ####
ggplot(het_long, aes(x = Year, y = Heterozygosity, color = Replicate)) +
  geom_line(linewidth = 1) +
  #scale_color_manual(values = c("He_MC1" = "#A0CBE8", "Ho_MC1" = "#4E79A7", "He_MC2" = "#F1CE63", "Ho_MC2" = "#B6992D", "He_MC3" = "#8CD17D", "Ho_MC3" = "#59A14F")) +
  theme_classic() +
  #theme(legend.position = c(0.9, 0.3)) +
  labs(x = "Year", y = "Heterozygosity") +
  ggtitle("Heterozygosity")+
  geom_vline(xintercept = 25, color = "black", linetype = "dashed")
```
