---
title: "null_visualizations_MC3"
author: "m quigg"
date: "2026-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Viola CDmetaPOP Simulation Visualization
This is to visualize the 3 MC replicates of the null model.  

## Set up
### Working Driectory and Library
```{r, include=FALSE}
setwd("C:/Users/mquig/Desktop/DED/LandGen_viola_sims")

library(tidyverse)
library(tools)
library(utils)
```

### Import Files
```{r}
#### MC1 ####
data_null_pop_MC1 = read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc0species0/summary_popAllTime.csv", stringsAsFactors = FALSE, sep = ",")

data_null_class_MC1=read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc0species0/summary_classAllTime.csv", stringsAsFactors = FALSE, sep = ",")


#### MC2 ####
data_null_pop_MC2 = read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc1species0/summary_popAllTime.csv", stringsAsFactors = FALSE, sep = ",")

data_null_class_MC2=read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc1species0/summary_classAllTime.csv", stringsAsFactors = FALSE, sep = ",")


#### MC3 ####
data_null_pop_MC3 = read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc2species0/summary_popAllTime.csv", stringsAsFactors = FALSE, sep = ",")

data_null_class_MC3=read.csv("C:/Users/mquig/Desktop/DED/LandGen_viola_sims/run0batch0mc2species0/summary_classAllTime.csv", stringsAsFactors = FALSE, sep = ",")
```

## Population Over Time
The goal is to make a graph that depicts the total population across all patches over time, for each of the MCs. It is a line graph with 3 lines, one for each replicate.
```{r}
#### Extract the data ####

### MC1 ###
pop_size_MC1=data_null_pop_MC1 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("MC1", paste0("patch_", 1:9)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("patch_"), as.numeric)) %>% 
  select(Year, MC1)

### MC2 ###
pop_size_MC2=data_null_pop_MC2 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("MC2", paste0("patch_", 1:9)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("patch_"), as.numeric)) %>% 
  select(Year, MC2)

### MC3 ###
pop_size_MC3=data_null_pop_MC3 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("MC3", paste0("patch_", 1:9)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("patch_"), as.numeric)) %>% 
  select(Year, MC3)


#### Make new dataframe ####
###combine dataframes
pop_size_MC=merge(merge(pop_size_MC1, pop_size_MC2, by="Year"), pop_size_MC3, by="Year")

###Convert to long format for graphing
pop_size_long=tidyr::pivot_longer(pop_size_MC, cols = starts_with("MC"), names_to = "MC", values_to = "Population Size")

###calculate CIs###
pop_summary=pop_size_long %>% 
  group_by(Year) %>% 
  summarize(n=length(`Population Size`),
            mean=mean(as.numeric(`Population Size`)),
            sd=sd(`Population Size`),
            se=sd/sqrt(n),
            lower.CI=mean-(1.96*se),
            upper.CI=mean+(1.96*se))

#### graph it! ####
ggplot(pop_summary, aes(x = Year, y = as.numeric(mean))) +
  geom_line(linewidth = 1, color = "#60b74d") +
  geom_ribbon(aes(ymin = lower.CI, ymax = upper.CI), alpha = 0.4, color = "#60b74d") +
  theme_classic() +
  labs(x = "Year", y = "Population") +
  ggtitle("Total Population")
```

## Population Size by Cohort
The goal here is to create 3 graphs (one per MC) that show the distribution of population size based on age classes.
```{r}
#### Extract the age data ####
### MC 1###
age_pop_MC1=data_null_class_MC1 %>% 
  select(Year, N_Initial_Class) %>% 
  separate(N_Initial_Class, into = c("MC1", paste0("age_", 0:24)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("age_"), as.numeric)) %>% 
  select(-MC1)

### MC 2###
age_pop_MC2=data_null_class_MC2 %>% 
  select(Year, N_Initial_Class) %>% 
  separate(N_Initial_Class, into = c("MC2", paste0("age_", 0:24)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("age_"), as.numeric)) %>% 
  select(-MC2)

### MC 3###
age_pop_MC3=data_null_class_MC3 %>% 
  select(Year, N_Initial_Class) %>% 
  separate(N_Initial_Class, into = c("MC3", paste0("age_", 0:24)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("age_"), as.numeric)) %>% 
  select(-MC3)

#### convert to long format ####
### MC1 ###
##pivot
long_age_MC1=tidyr::pivot_longer(age_pop_MC1, cols = starts_with("age_"), names_to = "Age", values_to = "Population")
##clean up columns
cohort_age_MC1=long_age_MC1 %>% 
  separate(Age, into = c("delete", "Age"), sep = "_") %>% 
  select(-delete)
cohort_age_MC1$Age=as.numeric(cohort_age_MC1$Age)


### MC2 ###
##pivot
long_age_MC2=tidyr::pivot_longer(age_pop_MC2, cols = starts_with("age_"), names_to = "Age", values_to = "Population")
##clean up columns
cohort_age_MC2=long_age_MC2 %>% 
  separate(Age, into = c("delete", "Age"), sep = "_") %>% 
  select(-delete) 
cohort_age_MC2$Age=as.numeric(cohort_age_MC2$Age)

### MC3 ###
##pivot
long_age_MC3=tidyr::pivot_longer(age_pop_MC3, cols = starts_with("age_"), names_to = "Age", values_to = "Population")
##clean up columns
cohort_age_MC3=long_age_MC3 %>% 
  separate(Age, into = c("delete", "Age"), sep = "_") %>% 
  select(-delete)
cohort_age_MC3$Age=as.numeric(cohort_age_MC3$Age)


#### group into cohorts ####
### MC1 ###
##organize into classes
cohort_MC1=cohort_age_MC1 %>% 
  mutate(
    cohort = case_when(
      Age <=5 ~ "Ages 0-5",
      Age >= 6 & Age <=10 ~ "Ages 6-10",
      Age >= 11 & Age <=15 ~ "Ages 11-15",
      Age >= 16 & Age <=20 ~ "Ages 16-20",
      Age >= 21 & Age <=25 ~ "Ages 21-25")) %>% 
  na.omit() %>% 
  select(-Age)
##summarize populations in classes
cohort_MC1_summary=cohort_MC1 %>% 
  group_by(Year, cohort) %>% 
  summarise(population_MC1=sum(Population))

### MC2 ###
##organize into classes
cohort_MC2=cohort_age_MC2 %>% 
  mutate(
    cohort = case_when(
      Age <=5 ~ "Ages 0-5",
      Age >= 6 & Age <=10 ~ "Ages 6-10",
      Age >= 11 & Age <=15 ~ "Ages 11-15",
      Age >= 16 & Age <=20 ~ "Ages 16-20",
      Age >= 21 & Age <=25 ~ "Ages 21-25")) %>% 
  na.omit() %>% 
  select(-Age)
##summarize populations in classes
cohort_MC2_summary=cohort_MC2 %>% 
  group_by(Year, cohort) %>% 
  summarise(population_MC2=sum(Population))

### MC3 ###
##organize into classes
cohort_MC3=cohort_age_MC3 %>% 
  mutate(
    cohort = case_when(
      Age <=5 ~ "Ages 0-5",
      Age >= 6 & Age <=10 ~ "Ages 6-10",
      Age >= 11 & Age <=15 ~ "Ages 11-15",
      Age >= 16 & Age <=20 ~ "Ages 16-20",
      Age >= 21 & Age <=25 ~ "Ages 21-25")) %>% 
  na.omit() %>% 
  select(-Age)
##summarize populations in classes
cohort_MC3_summary=cohort_MC3 %>% 
  group_by(Year, cohort) %>% 
  summarise(population_MC3=sum(Population))

#### graph it ####
### MC1 ###
ggplot(cohort_MC1_summary, aes(x = Year, y = population_MC1, color = cohort)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "Cohort") +
  ggtitle("MC1")

### MC2 ###
ggplot(cohort_MC2_summary, aes(x = Year, y = population_MC2, color = cohort)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "Cohort") +
  ggtitle("MC2")

### MC3 ###
ggplot(cohort_MC3_summary, aes(x = Year, y = population_MC3, color = cohort)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "Cohort") +
  ggtitle("MC3")
```
### Cohort with Confidence Intervals
The goal is to make one graph with all of the cohort pop sizes on one graph with confidence intervals.
```{r}
#### Extract and process all MC replicates ####

process_mc <- function(data, mc_num) {
  data %>%
    select(Year, N_Initial_Class) %>%
    separate(N_Initial_Class, 
             into = c("drop_col", paste0("age_", 0:24)), 
             sep = "\\|", extra = "drop") %>%
    mutate(across(starts_with("age_"), as.numeric)) %>%
    select(-drop_col) %>%
    pivot_longer(cols = starts_with("age_"), 
                 names_to = "Age", 
                 values_to = "Population") %>%
    separate(Age, into = c("delete", "Age"), sep = "_") %>%
    select(-delete) %>%
    mutate(
      Age = as.numeric(Age),
      cohort = case_when(
        Age <= 5              ~ "Ages 0-5",
        Age >= 6  & Age <= 10 ~ "Ages 6-10",
        Age >= 11 & Age <= 15 ~ "Ages 11-15",
        Age >= 16 & Age <= 20 ~ "Ages 16-20",
        Age >= 21 & Age <= 25 ~ "Ages 21-25"),
      MC = paste0("MC", mc_num)) %>%
    na.omit() %>%
    select(-Age) %>%
    group_by(Year, cohort, MC) %>%
    summarise(Population = sum(Population), .groups = "drop")
}

# Process each replicate
mc1 <- process_mc(data_null_class_MC1, 1)
mc2 <- process_mc(data_null_class_MC2, 2)
mc3 <- process_mc(data_null_class_MC3, 3)

#### Combine and calculate mean + 95% CI ####

cohort_combined <- bind_rows(mc1, mc2, mc3) %>%
  group_by(Year, cohort) %>%
  summarise(
    mean_pop = mean(Population),
    sd_pop   = sd(Population),
    n        = n(),
    se       = sd_pop / sqrt(n),
    ci_lower = mean_pop - qt(0.95, df = n - 1) * se,
    ci_upper = mean_pop + qt(0.95, df = n - 1) * se,
    .groups = "drop"
  )

#### Plot ####

ggplot(cohort_combined, aes(x = Year, y = mean_pop, color = cohort, fill = cohort)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "Cohort", fill = "Cohort",
       title = "Mean Population by Cohort (± 95% CI across MC replicates)")

```

## Population by Patch
The goal here is to create 3 graphs (one for each MC) depicting the population by patch over time.
```{r}
#### MC1 ####
###extract pop data by patch
pop_patch_MC1=data_null_pop_MC1 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("patch_"), as.numeric))
 
###pivot
long_pop_patch_MC1=tidyr::pivot_longer(pop_patch_MC1, cols = starts_with("patch_"), names_to = "Patch", values_to = "Population")

###graph it
ggplot(long_pop_patch_MC1, aes(x = Year, y = Population, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population Size", color = "Patch") +
  ggtitle("MC1") 
  #+ geom_vline(xintercept = 25, color = "black", linetype = "dashed")
  
#### MC2 ####
###extract pop data by patch
pop_patch_MC2=data_null_pop_MC2 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("patch_"), as.numeric))
 
###pivot
long_pop_patch_MC2=tidyr::pivot_longer(pop_patch_MC2, cols = starts_with("patch_"), names_to = "Patch", values_to = "Population")

###graph it
ggplot(long_pop_patch_MC2, aes(x = Year, y = Population, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population Size", color = "Patch") +
  ggtitle("MC2")   
  #+ geom_vline(xintercept = 25, color = "black", linetype = "dashed")
  
#### MC3 ####
###extract pop data by patch
pop_patch_MC3=data_null_pop_MC3 %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("patch_"), as.numeric))
 
###pivot
long_pop_patch_MC3=tidyr::pivot_longer(pop_patch_MC3, cols = starts_with("patch_"), names_to = "Patch", values_to = "Population")

###graph it
ggplot(long_pop_patch_MC3, aes(x = Year, y = Population, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population Size", color = "Patch") +
  ggtitle("MC3")   
  #+ geom_vline(xintercept = 25, color = "black", linetype = "dashed")
```
### Patch population with CI
```{r}
#### Process all MC replicates ####

process_mc_patch <- function(data, mc_num) {
  data %>%
    select(Year, N_Initial) %>%
    separate(N_Initial,
             into = c("all", paste0("patch_", 1:9)),
             sep = "\\|", extra = "drop") %>%
    mutate(across(starts_with("patch_"), as.numeric)) %>%
    select(-all) %>%
    pivot_longer(cols = starts_with("patch_"),
                 names_to = "Patch",
                 values_to = "Population") %>%
    mutate(MC = paste0("MC", mc_num))
}

mc1_patch <- process_mc_patch(data_null_pop_MC1, 1)
mc2_patch <- process_mc_patch(data_null_pop_MC2, 2)
mc3_patch <- process_mc_patch(data_null_pop_MC3, 3)

#### Combine and calculate mean + 95% CI ####

patch_combined <- bind_rows(mc1_patch, mc2_patch, mc3_patch) %>%
  group_by(Year, Patch) %>%
  summarise(
    mean_pop = mean(Population),
    sd_pop   = sd(Population),
    n        = n(),
    se       = sd_pop / sqrt(n),
    ci_lower = mean_pop - qt(0.975, df = n - 1) * se,
    ci_upper = mean_pop + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )

#### Plot ####

ggplot(patch_combined, aes(x = Year, y = mean_pop, color = Patch, fill = Patch)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population Size", color = "Patch", fill = "Patch",
       title = "Mean Population by Patch (± 95% CI across MC replicates)")
  #+ geom_vline(xintercept = 25, color = "black", linetype = "dashed")
```


## Heterozygosity
I want one graph with observed and expected heterozygosity for each of the MC replicates.
```{r}
#### extract data ####
### MC1 ###
het_MC1=data_null_pop_MC1 %>% 
  select(Year, Ho, He) %>% 
  separate(Ho, into = c("Ho_MC1", paste0("Ho_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  separate(He, into = c("He_MC1", paste0("He_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("Ho"), as.numeric)) %>% 
  mutate(across(starts_with("He"), as.numeric)) %>% 
  select(Year, Ho_MC1, He_MC1)

### MC2 ###
het_MC2=data_null_pop_MC2 %>% 
  select(Year, Ho, He) %>% 
  separate(Ho, into = c("Ho_MC2", paste0("Ho_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  separate(He, into = c("He_MC2", paste0("He_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("Ho"), as.numeric)) %>% 
  mutate(across(starts_with("He"), as.numeric)) %>% 
  select(Year, Ho_MC2, He_MC2)

### MC3 ###
het_MC3=data_null_pop_MC3 %>% 
  select(Year, Ho, He) %>% 
  separate(Ho, into = c("Ho_MC3", paste0("Ho_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  separate(He, into = c("He_MC3", paste0("He_patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("Ho"), as.numeric)) %>% 
  mutate(across(starts_with("He"), as.numeric)) %>% 
  select(Year, Ho_MC3, He_MC3)

#### Make new dataframe ####
###combine dataframes
het_MC=merge(merge(het_MC1, het_MC2, by="Year"), het_MC3, by="Year")

###Convert to long format for graphing
het_long=tidyr::pivot_longer(het_MC, cols = starts_with("H"), names_to = "Replicate", values_to = "Heterozygosity")

#### graph it! ####
ggplot(het_long, aes(x = Year, y = Heterozygosity, color = Replicate)) +
  geom_line(linewidth = 1) +
  #scale_color_manual(values = c("He_MC1" = "#A0CBE8", "Ho_MC1" = "#4E79A7", "He_MC2" = "#F1CE63", "Ho_MC2" = "#B6992D", "He_MC3" = "#8CD17D", "Ho_MC3" = "#59A14F")) +
  theme_classic() +
  #theme(legend.position = c(0.9, 0.3)) +
  labs(x = "Year", y = "Heterozygosity") +
  ggtitle("Heterozygosity")+
  geom_vline(xintercept = 25, color = "black", linetype = "dashed")
```

###One graph with CIs
```{r}
#### Process all MC replicates ####

process_mc_het <- function(data, mc_num) {
  data %>%
    select(Year, Ho, He) %>%
    separate(Ho, into = c("Ho_val", paste0("Ho_patch_", 1:9)), sep = "\\|", extra = "drop") %>%
    separate(He, into = c("He_val", paste0("He_patch_", 1:9)), sep = "\\|", extra = "drop") %>%
    mutate(across(starts_with("Ho"), as.numeric),
           across(starts_with("He"), as.numeric)) %>%
    select(Year, Ho = Ho_val, He = He_val) %>%
    pivot_longer(cols = c(Ho, He), names_to = "Metric", values_to = "Heterozygosity") %>%
    mutate(MC = paste0("MC", mc_num))
}

mc1_het <- process_mc_het(data_null_pop_MC1, 1)
mc2_het <- process_mc_het(data_null_pop_MC2, 2)
mc3_het <- process_mc_het(data_null_pop_MC3, 3)

#### Combine and calculate mean + 95% CI ####

het_combined <- bind_rows(mc1_het, mc2_het, mc3_het) %>%
  group_by(Year, Metric) %>%
  summarise(
    mean_het = mean(Heterozygosity),
    sd_het   = sd(Heterozygosity),
    n        = n(),
    se       = sd_het / sqrt(n),
    ci_lower = mean_het - qt(0.975, df = n - 1) * se,
    ci_upper = mean_het + qt(0.975, df = n - 1) * se,
    .groups = "drop"
  )

#### Plot ####

ggplot(het_combined, aes(x = Year, y = mean_het, color = Metric, fill = Metric)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Heterozygosity", color = "Metric", fill = "Metric",
       title = "Heterozygosity (± 95% CI across MC replicates)") +
  geom_vline(xintercept = 25, color = "black", linetype = "dashed")
```
