---
title: "viola_output"
author: "m quigg"
date: "2026-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Viola CDmetaPOP Simulation Visualization

Basic commands and prep
```{r, include=FALSE}
setwd("C:/Users/mquig/Desktop/DED/LandGen_viola_sims")

library(tidyverse)
library(tools)
library(utils)
```

## Null Model
This model does not incorporate herbivory or climate change.

###Prep
```{r}
###import file
data_null_pop = read.csv("summary_popAllTime_null.csv", stringsAsFactors = FALSE, sep = ",")
data_null_class=read.csv("summary_classAllTime_null.csv", stringsAsFactors = FALSE, sep = ",")
```


### Total population over time
```{r}
###set up the data
data_null_pop <- data_null_pop %>%
  mutate(N_Initial = str_replace(N_Initial, "\\|$", ""))
n_cols <- max(str_count(data_null_pop$N_Initial, "\\|"))
patch_names <- c("patch_total", paste0("patch_", 1:n_cols))
age_data_null <- data_null_pop %>%
  separate(
    N_Initial,
    into = patch_names,
    sep = "\\|",
    convert = TRUE,
    fill = "right")
age_data_null_patches=age_data_null %>% 
  select(Year, patch_total)

###graph it
ggplot(age_data_null_patches, aes(x = Year, y = patch_total)) +
    geom_line(color = "steelblue", linewidth = 0.8) +
    labs(title = "Population Size (null)", y = "Population Size", x = "Year") +
  theme_classic()
```
This is good and accurate.

```{r}
####pop size by age class (cohorts)####
###extract the data###
age_data_null=read.table(text = data_null_class$N_Initial_Class, sep = "|")
colnames(age_data_null)=paste0("Age_", 0:(ncol(age_data_null)-1))
age_data_null$Year=data_null_class$Year

###pivot the dataframe
long_data_null=tidyr::pivot_longer(age_data_null, cols = starts_with("Age_"), names_to = "Ages", values_to = "Count")

cohort_data_null=long_data_null %>% 
  separate(Ages, into = c("delete", "Age"), sep = "_")
###make proper type
cohort_data_null$Age=as.numeric(cohort_data_null$Age)

###organize into classes
cohort_data_null=cohort_data_null %>% 
  mutate(
    cohort = case_when(
      Age <=5 ~ "Ages 0-5",
      Age >= 6 & Age <=10 ~ "Ages 6-10",
      Age >= 11 & Age <=15 ~ "Ages 11-15",
      Age >= 16 & Age <=20 ~ "Ages 16-20",
      Age >= 21 & Age <=25 ~ "Ages 21-25")) %>% 
  na.omit() %>% 
  select(-delete, -Age)

###summarize populations in classes
cohort_data2_null=cohort_data_null %>% 
  group_by(Year, cohort)
cohort_data3_table_null= cohort_data2_null %>% 
  summarise(total_count= sum(Count))

###graph it  
ggplot(cohort_data3_table_null, aes(x = Year, y = total_count, color = cohort)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population", color = "Cohort") +
  ggtitle("Population by Cohort (null)")
```
###Population by Patch over time
NOw I want to take a look at the population of each patch over time.
```{r}
###extract pop data by patch
pop_patch_null=data_null_pop %>% 
  select(Year, N_Initial) %>% 
  separate(N_Initial, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("patch_"), as.numeric))
 
###pivot
long_pop_patch_null=tidyr::pivot_longer(pop_patch_null, cols = starts_with("patch_"), names_to = "Patch", values_to = "Population")

###graph it
ggplot(long_pop_patch_null, aes(x = Year, y = Population, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Population Size", color = "Patch") +
  ggtitle("Population of Each Patch Over Time") +
  geom_vline(xintercept = 25, color = "black", linetype = "dashed")
```




###Heterozygosity over time
Heterozygosity information can be found in the pop file.
```{r}
####observed heterozygosity####
###extract observed heterozygosity data
obs_het_null=data_null_pop %>% 
  select(Year, Ho) %>% 
  separate(Ho, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop") %>% 
  mutate(across(starts_with("patch_"), as.numeric))
#%>% 
#  rowwise() %>%
#  mutate(mean = mean(c_across(starts_with("patch")), na.rm = TRUE)) %>%
#  ungroup()

###pivot
long_obs_het_null=tidyr::pivot_longer(obs_het_null, cols = starts_with("patch_"), names_to = "Patch", values_to = "Ho")

###graph it
ggplot(long_obs_het_null, aes(x = Year, y = Ho, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Ho", color = "Patch") +
  ggtitle("Observed Heterozygosity Over Time") +
  geom_vline(xintercept = 25, color = "black", linetype = "dashed")

####expected heterozygosity####
###extract observed heterozygosity data
ex_het_null=data_null_pop %>% 
  select(Year, He) %>% 
  separate(He, into = c("all", paste0("patch_", 1:9)), sep = "\\|", extra = "drop")  %>% 
    mutate(across(starts_with("patch_"), as.numeric))
#  rowwise() %>%
#  mutate(mean = mean(c_across(starts_with("patch")), na.rm = TRUE)) %>%
#  ungroup()

###pivot
long_ex_het_null=
  tidyr::pivot_longer(ex_het_null, cols = starts_with("patch_"), names_to = "Patch", values_to = "He")

###graph it
ggplot(long_ex_het_null, aes(x = Year, y = He, color = Patch)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "He", color = "Patch") +
  ggtitle("Expected Heterozygosity Over Time") +
  geom_vline(xintercept = 25, color = "black", linetype = "dashed")

####both####
###extract mean values across patches
mean_obs_het_null=obs_het_null %>% 
  select(Year, all) %>% 
  rename(Ho=all)
mean_ex_het_null=ex_het_null %>% 
  select(Year, all) %>% 
  rename(He=all)
###conbine dataframe
mean_het_null=left_join(mean_obs_het_null, mean_ex_het_null, by = "Year")
###pivot
long_mean_het_null=tidyr::pivot_longer(mean_het_null, cols = c("Ho", "He"), names_to = "Heterozygosity", values_to = "mean_values")
###graph it
ggplot(long_mean_het_null, aes(x = Year, y = as.numeric(mean_values), color = Heterozygosity)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Heterozygosity") +
  ggtitle("Heterozygosity Over Time Across all Patches")+
  geom_vline(xintercept = 25, color = "black", linetype = "dashed")
```


## Herbivory
This is to visualize the scenario with the output of herbivory simulations.
###Prep
```{r}
###import file
data_herb_pop = read.csv("summary_popAllTime_herb.csv", stringsAsFactors = FALSE, sep = ",")
data_herb_class=read.csv("summary_classAllTime_herb.csv", stringsAsFactors = FALSE, sep = ",")
```

*went extinct after 48 years

### Total population over time
```{r}
###set up the data
data_herb_pop <- data_herb_pop %>%
  mutate(N_Initial = str_replace(N_Initial, "\\|$", ""))
n_cols <- max(str_count(data_herb_pop$N_Initial, "\\|"))
patch_names <- c("patch_total", paste0("patch_", 1:n_cols))
age_data_herb <- data_herb_pop %>%
  separate(
    N_Initial,
    into = patch_names,
    sep = "\\|",
    convert = TRUE,
    fill = "right")
age_data_herb_patches=age_data_herb %>% 
  select(Year, patch_total)

###graph it
ggplot(age_data_herb_patches, aes(x = Year, y = patch_total)) +
    geom_line(color = "steelblue", linewidth = 0.8) +
    labs(title = "Population Size (herbivory)", y = "Population Size", x = "Year") +
  theme_classic() +
  geom_vline(xintercept = 48, color = "red")
```
This is good and accurate.

```{r}
####pop size by age class (cohorts)####
###extract the data###
age_data_herb=read.table(text = data_herb_class$N_Initial_Class, sep = "|")
colnames(age_data_herb)=paste0("Age_", 0:(ncol(age_data_herb)-1))
age_data_herb$Year=data_herb_class$Year

###pivot the dataframe
long_data_herb=tidyr::pivot_longer(age_data_herb, cols = starts_with("Age_"), names_to = "Ages", values_to = "Count")

cohort_data_herb=long_data_herb %>% 
  separate(Ages, into = c("delete", "Age"), sep = "_")
###make proper type
cohort_data_herb$Age=as.numeric(cohort_data_herb$Age)

###organize into classes
cohort_data_herb=cohort_data_herb %>% 
  mutate(
    cohort = case_when(
      Age <=5 ~ "Ages 0-5",
      Age >= 6 & Age <=10 ~ "Ages 6-10",
      Age >= 11 & Age <=15 ~ "Ages 11-15",
      Age >= 16 & Age <=20 ~ "Ages 16-20",
      Age >= 21 & Age <=25 ~ "Ages 21-25")) %>% 
  na.omit() %>% 
  select(-delete, -Age)

###summarize populations in classes
cohort_data2_herb=cohort_data_herb %>% 
  group_by(Year, cohort)
cohort_data3_table_herb= cohort_data2_herb %>% 
  summarise(total_count= sum(Count))

###graph it  
ggplot(cohort_data3_table_herb, aes(x = Year, y = total_count, color = cohort)) +
  geom_line(linewidth = 1) +
  theme_classic() +
  labs(x = "Year", y = "Count", color = "Cohort") +
  ggtitle("Population by Cohort (herbivory)")
```
Yay it worked!
